---
title: "MOMS-PI"
# author: "Author's Name"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: cerulean
    toc: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

```{r}
library(readr)
library(DataExplorer)
library(reshape2)
```


```{r settings}
# Data folder
dataDir <- "/Users/mdozmorov/Documents/Data/GitHub/HMP2/MOMS-PI/"
# Input files
fileNameIn1  <- "MOMS-PI_16S_STIRRUPS_profiles_POP1.txt"
fileNameIn2  <- "MOMS-PI_CytokineProfiles_POP1.txt"
fileNameIn3  <- "MOMS-PI_LipidomicsProfiles_POP1.txt"
fileNameIn4  <- "MOMS-PI_subject_kit_sample_mapping_POP1.txt"
# Output files
fileNameOut1 <- "results/report_momspi.html"
```

# Exploratory data analysis

Data source: http://vmc.vcu.edu/resources/momspi. Download with `scripts/download_moms-pi.sh`

## `r fileNameIn1`

```{r}
# mtx1 <- read_tsv(paste0(dataDir, fileNameIn1), col_names = TRUE)
mtx1 <- read.table(paste0(dataDir, fileNameIn1), sep = "\t", header = TRUE)
dim(mtx1)
head(mtx1) %>% pander
# Fix numbers
mtx1$No_of_Reads          <- as.numeric(mtx1$No_of_Reads)
mtx1$Percentage_Abundance <- as.numeric(mtx1$Percentage_Abundance)
mtx1$Avg._Score           <- as.numeric(mtx1$Avg._Score)
```

**Questions**

- What are the "Threshold.Status", "No_of_Reads", "Percentage_Abundance", "Avg._Score" columns?

```{r eval=FALSE}
### Standard report
create_report(mtx1, output_file = fileNameOut1, output_dir = "results/momspi")
```

**Manual EDA**

```{r}
print("How many unique samples?")
mtx1$SampleID %>% unique() %>% length() # 69
print("How many unique taxa?")
mtx1$Taxa %>% unique() %>% length() # 299
print("How many unique Threshold.Status?")
table(mtx1$Threshold.Status)
print("What is the distribution of the No_of_Reads?")
plot_density(mtx1$No_of_Reads %>% log2, title = "log2 No_of_Reads density plot")
summary(mtx1$No_of_Reads)
print("What is the distribution of the Percentage_Abundance?")
plot_density(mtx1$Percentage_Abundance %>% log2, title = "log2 Percentage_Abundance density plot")
summary(mtx1$Percentage_Abundance)
print("What is the distribution of the Avg._Score?")
plot_density((mtx1$Avg._Score + 1) %>% log2, title = "log2 Avg._Score density plot")
summary(mtx1$Avg._Score)
```

**Questions**

- No_of_Reads, Percentage_Abundance, Avg._Score have outlies - is that normal?

Convert to wide format

```{r}
mtx1_wide <- dcast(mtx1[, c("SampleID", "Taxa", "Percentage_Abundance")], SampleID ~ Taxa, value.var = "Percentage_Abundance", fun.aggregate = sum)
dim(mtx1_wide)
mtx1_wide[1:15, 1:5] %>% pander
```

**Questions**

- Many zeros (missing values?) - is that normal?

What are the most abundant strains?

```{r}
mtx1_wide %>% select(-SampleID) %>% apply(., 2, sum) %>% sort(., decreasing = TRUE) %>% head
```

## `r fileNameIn2`

```{r}
mtx2 <- read.table(paste0(dataDir, fileNameIn2), sep = "\t", header = TRUE, fill = TRUE)
dim(mtx2)
head(mtx2) %>% pander
# Fix numbers
```

## `r fileNameIn3`

```{r}
mtx3 <- read.table(paste0(dataDir, fileNameIn3), sep = "\t", header = TRUE, fill = TRUE)
dim(mtx3)
head(mtx3) %>% pander
# Fix numbers
```

## `r fileNameIn4`

```{r}
mtx4 <- read.table(paste0(dataDir, fileNameIn4), sep = "\t", header = TRUE, fill = TRUE)
dim(mtx4)
head(mtx4) %>% pander
# Fix numbers
```

**Manual EDA**

```{r}
print("How many unique SubjectID?")
mtx4$SubjectID %>% unique() %>% length()
print("How many unique SampleType?")
table(mtx4$SampleType)
print("How many BodySite?")
table(mtx4$BodySite)
print("How many VisitNumber?")
table(mtx4$VisitNumber)
```

**Questions**

- 74 unique subjects, but should be 69?
- What is "mtx4$SampleType"?