---
title: "Exploratory data analysis of cytokine files"
# author: "Author's Name"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: cerulean
    toc: yes
---


```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=T, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

```{r}
library(readr)
library(magrittr)
library(biomformat)
library(ggplot2)
library(DataExplorer)
library(reshape2)
library(openxlsx)
library(MDmisc)
```

# Settings

```{r settings}
# Big Data folder
dataDir      <- "/Users/mdozmorov/Documents/Data/GitHub/HMP2/"

# Input files
fileNameIn1  <- "mtx_cytokine_data_all_ID_merged.xlsx"  # All data from cytokine files, merged by ID
fileNameIn2  <- "data/hmp_cart_metadata_1e3cf1e9c1.tsv" # Metadata
fileNameIn3  <- "data/hmp_cart_b10350603.tsv" # Manifest
# Output files
fileNameOut1 <- "data/hmp2_cytokines_momspi.rda"

# Misc
max_samples  <- 500 # For some time-consuming operations, process that many randomly selected samples
```

# Load data

Data source: https://www.hmpdacc.org/hmp/, https://portal.hmpdacc.org/ - data portal. [Samples/Studies: MOMS-PI, Files/Matrix Type: "host_cytokine" - selects 872 files](https://portal.hmpdacc.org/search/f?filters=%7B%22op%22:%22and%22,%22content%22:%5B%7B%22op%22:%22in%22,%22content%22:%7B%22field%22:%22cases.study_name%22,%22value%22:%5B%22MOMS-PI%22%5D%7D%7D,%7B%22op%22:%22in%22,%22content%22:%7B%22field%22:%22files.file_matrix_type%22,%22value%22:%5B%22host_cytokine%22%5D%7D%7D%5D%7D&facetTab=files&pagination=%7B%22files%22:%7B%22from%22:0,%22size%22:20,%22sort%22:%22file_name.raw:asc%22%7D%7D). Download with `scripts/ascp-commands_biom_host_cytokine.sh`

```{r}
# All data
mtx_data_all_df <- read.xlsx(xlsxFile = paste0(dataDir, fileNameIn1), sheet = "Cytokine")
# Clean columns from ".otu_table.biom" suffix
colnames(mtx_data_all_df) <- sub(pattern = ".(pg.cytokine/mg.protein)", replacement = "", colnames(mtx_data_all_df), fixed = TRUE)
# Make IDs numeric
# mtx_data_all_df$ID <- as.numeric(mtx_data_all_df$ID)
```

There are NAs and values like "< OOR". The latter are replaced by 0.

```{r}
mtx_data_all_df[mtx_data_all_df == "OOR <"] <- 0 # Replace "OOR <" by 0
# Make numeric. Danger zone
mtx_data_all_df[, colnames(mtx_data_all_df) != "Cytokine"] <- lapply(mtx_data_all_df[, colnames(mtx_data_all_df) != "Cytokine"],
                                                                     function(x) {as.numeric(x)})
# Check that the data is numeric
# sapply(mtx_data_all_df[, colnames(mtx_data_all_df) != "Cytokine"], class)
```


```{r}
# Metadata
mtx_metadata <- read_tsv(fileNameIn2)
# Manifest
mtx_manifest <- read_tsv(fileNameIn3)
# Select cpecific columns from the manifest
mtx_manifest_selected <- data.frame(sample_id = mtx_manifest$sample_id, file = basename(mtx_manifest$urls))
# Clean columns from ".otu_table.biom" suffix
mtx_manifest_selected$file <- sub(pattern = ".CytokineProfile.txt", replacement = "", mtx_manifest_selected$file, fixed = TRUE)
# Attach file names to Metadata
mtx_metadata <- left_join(mtx_metadata, mtx_manifest_selected, by = c("sample_id"))
# Subset to the files in mtx_data_all_df
mtx_metadata <- mtx_metadata[mtx_metadata$file %in% colnames(mtx_data_all_df %>% dplyr::select(-Cytokine)), ]
# Make the same order as in mtx_data_all_df
mtx_metadata <- mtx_metadata[match(colnames(mtx_data_all_df %>% dplyr::select(-Cytokine)), mtx_metadata$file), ]
# Sanity check, should be TRUE
# all_equal(mtx_metadata$file, colnames(mtx_data_all_df %>% dplyr::select(-Cytokine)))
```


## Cytokine EDA

Data dimensions: `r dim(mtx_data_all_df)`

```{r}
mtx_data_all_df[1:5, 1:5]
```

\pagebreak

### Count distribution for `r max_samples` randomly selected samples

```{r fig.height=4}
set.seed(1)
# Prepare data for multiple histograms
data <- mtx_data_all_df[, sample(2:ncol(mtx_data_all_df), size = max_samples)] %>% melt # Randomly select max_samples
data <- data[ complete.cases(data), ] # Remove NAs
print("Summary of count data")
summary(data$value)
data$value <- log2(data$value + 1) # log-transform
# Overlaying density curves
ggplot(data, aes(x = value, fill = variable)) + 
  geom_density(alpha = 0.25) + 
  guides(fill = FALSE) +
  ggtitle("Count distribution") +
  xlab("log2 non-NA counts") +
  ylab("Density")
```

**Question:** There are outlier counts. With median/mean counts equal to 24/1363, respectively, we have  counts as high as 1008760.

\pagebreak

### Distribution of sample medians for all samples

```{r fig.height=4}
mtx_medians <- sapply(mtx_data_all_df %>% dplyr::select(-Cytokine), function(x) median(x, na.rm = TRUE))
summary(mtx_medians)
plot_density(mtx_medians)
```

**Question:** The range of medians vary across samples. Most samples have median counts ~25, but some have as high as 563.

### Boxplots of low/high median samples

We see how different are count distributions between low/high median samples

```{r fig.height=4}
# Sample indices with high and low median counts
index_high <- which(mtx_medians > quantile(mtx_medians, .99))
index_low  <- which(mtx_medians < quantile(mtx_medians, .17))
# Randomly subset ten of them from each group
index_high <- index_high[sample(x = 1:length(index_high), size = 9)]
index_low  <- index_low[sample( x = 1:length(index_low) , size = 9)]
# Get data for plotting
mtx_to_plot <- cbind(mtx_data_all_df[, index_low + 1], mtx_data_all_df[, index_high + 1]) # +1 because of "ID" column still present
mtx_to_plot <- mtx_to_plot %>% melt()
# Actually plot
ggplot(mtx_to_plot, aes(x = variable, y = value)) +
  geom_boxplot() +
  ylim(0, 100) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("Samples") + ylab("Counts") + ggtitle("Boxplots of low/high median samples")
```

\pagebreak

### Sparsity of samples (across columns)

```{r fig.height=3.4}
dim_rows <- nrow(mtx_data_all_df) # Total number of rows
# Percent of rows with NAs, per sample
pct_nas  <- sapply(mtx_data_all_df %>% dplyr::select(-Cytokine), function(x) sum(is.na(x)) / dim_rows) %>% unname
# Density histogram
data.frame(pct_nas) %>% plot_density(., title = "Sparsity distribution across samples")
# Summary statistics
summary(pct_nas)
```

**Conclusion:** Sparsity is present, but not bad

```{r}
# Add sample names to percent NAs
names(pct_nas) <- colnames(mtx_data_all_df %>% dplyr::select(-Cytokine))
pct_nas <- sort(pct_nas) # Sort from low to high

print("Top 10 least sparse samples:")
pct_nas_least <- pct_nas[1:10]
kable(mtx_metadata[mtx_metadata$file %in% names(pct_nas_least), c("file", "sample_body_site", "visit_number")])

print("Top 10 most sparse samples:")
pct_nas_most <- pct_nas[(length(pct_nas) - 10):length(pct_nas)]
kable(mtx_metadata[mtx_metadata$file %in% names(pct_nas_most), c("file", "sample_body_site", "visit_number")])
```

**Observation:** Samples from buccal mucosa may be least sparse. Samples from vagina and first visits may be most sparse.

\pagebreak

### Sparsity of counts (across rows)

```{r fig.height=3.5}
dim_cols <- ncol(mtx_data_all_df %>% dplyr::select(-Cytokine)) # Total number of columns
# Percent of rows with NAs, per sample
pct_nas  <- apply(mtx_data_all_df %>% dplyr::select(-Cytokine), 1, function(x) sum(is.na(x)) / dim_cols) %>% unname
# Density histogram
data.frame(pct_nas) %>% plot_density(., title = "Sparsity distribution across counts")
# Summary statistics
summary(pct_nas)
```

**Question:** The data across counts is OK except one measure that is very sparse.

```{r}
# Add sample names to percent NAs
names(pct_nas) <- mtx_data_all_df$Cytokine
pct_nas <- sort(pct_nas) # Sort from low to high

print("Top 5 least sparse counts:")
pct_nas_least <- pct_nas[1:5]
kable(pct_nas_least)

print("Top 5 most sparse counts:")
pct_nas_most <- pct_nas[(length(pct_nas) - 5):length(pct_nas)]
kable(pct_nas_most)
```

**Observation:** Counts for IL-5 and IL-2 cytokines are very sparse.

\pagebreak

## Metadata EDA

```{r echo=TRUE}
# Quick EDA
table(mtx_metadata$subject_gender)   %>% sort(., decreasing = TRUE)
table(mtx_metadata$subject_race)     %>% sort(., decreasing = TRUE)
table(mtx_metadata$study_full_name)  %>% sort(., decreasing = TRUE)
table(mtx_metadata$project_name)     %>% sort(., decreasing = TRUE)
```

### sample_body_site

```{r fig.height=4}
plot_bar(mtx_metadata$sample_body_site)
table(mtx_metadata$sample_body_site) %>% sort(., decreasing = TRUE)
```

### visit_number

```{r fig.height=4}
plot_density(mtx_metadata$visit_number)
table(mtx_metadata$visit_number)    %>% sort(., decreasing = TRUE)
```

```{r}
print(paste0("How many total samples: "     , ncol(mtx_data_all_df %>% dplyr::select(-Cytokine)) ))
print(paste0("How many samples at visit 1: ", sum(mtx_metadata$visit_number == 1) ))
```

### Number of subjects with vaginal samples

```{r}
dim(mtx_metadata)
mtx2_vaginal <- subset(mtx_metadata, sample_body_site == "vagina")
length(unique(mtx_metadata$subject_id))
nvisits <- aggregate(visit_number ~ subject_id,  data = mtx2_vaginal, length)
table(nvisits$visit_number)
```

## Save the results

```{r}
# Cytokine data
mtx_cytokine_data <- mtx_data_all_df
# Sanity check for file names in "mtx_data" are the same as in "mtx_metadata"
# all.equal(mtx_metadata$file, colnames(mtx_data)[grep("EP", colnames(mtx_data))])
save(list = c("mtx_cytokine_data", "mtx_metadata"), file = fileNameOut1)
# load(file = fileNameOut1)
```

