---
output: pdf_document
---

```{r}
library(phyloseq)
library(biomformat)
# Data folder, adjust as needed
dataDir <- "/Users/mdozmorov/Documents/Data/GitHub/HMP2/ptb_nonreadable/"
# Use phyloseq or low-level reading of BIOM files
phyloseq_import_biom <- FALSE # If FALSE, `read_biom` will be used
# All nonreadable files
files <- list.files(dataDir, pattern = "biom", full.names = TRUE)
length(files) 
```

```{r}
my_import_biom <- function(file, phyloseq_import_biom = TRUE) {
  if (phyloseq_import_biom) { # phyloseq way
    mtx_biom <- import_biom(file) %>% otu_table() %>% as.data.frame() 
    mtx_biom <- data.frame(ID = rownames(mtx_biom), mtx_biom)
  } else { # bioformat way
    mtx_biom <- read_biom(file)
    mtx_biom_data <- biom_data(mtx_biom, parallel = TRUE) # Extract data one time 
    mtx_biom <- data.frame(ID = names(mtx_biom_data), mtx_biom_data)
  }
  return(mtx_biom)
}
```


```{r}
# Test any file
mtx_biom <- my_import_biom(files[1], phyloseq_import_biom = FALSE) # biomformat way
```

Error in read_biom(file) : Both attempts to read input file:
/Users/mdozmorov/Documents/Data/GitHub/HMP2/ptb_nonreadable//EP016061_K20_MV1D.otu_table.biom
either as JSON (BIOM-v1) or HDF5 (BIOM-v2).
Check file path, file name, file itself, then try again.
In addition: Warning message:
In strsplit(conditionMessage(e), "\n") :

```{r}
mtx_biom <- my_import_biom(files[1], phyloseq_import_biom = TRUE) # phyloseq way
```

Error in read_biom(biom_file = BIOMfilename) : 
  Both attempts to read input file:
/Users/mdozmorov/Documents/Data/GitHub/HMP2/ptb_nonreadable//EP016061_K20_MV1D.otu_table.biom
either as JSON (BIOM-v1) or HDF5 (BIOM-v2).
Check file path, file name, file itself, then try again.
In addition: Warning message:
In strsplit(conditionMessage(e), "\n") :

```{r eval=FALSE}
# biomformat way
mtx_biom <- read_biom(files[1]) 
# Testing biomformat functions
biom_data(mtx_biom) %>% as.data.frame()
biom_shape(mtx_biom)
colnames(mtx_biom)
rownames(mtx_biom)
header(mtx_biom)
```

