---
title: "16S and cytokines combination"
output: pdf_document
---

Check: Evaluation of integrative clustering methods for the analysis of multi-omics data
https://academic.oup.com/bib/advance-article-abstract/doi/10.1093/bib/bbz015/5316049?redirectedFrom=fulltext

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning= FALSE)
```

```{r, echo = FALSE}
#token 35f0903098e0c9ef30faf7514e382e6bbd5c7179
#BiocManager::install("HMP2Data")

#devtools::install_github("jstansfield0/HMP2Data", auth_token = "35f0903098e0c9ef30faf7514e382e6bbd5c7179")

library(HMP2Data)
library(phyloseq)
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(dplyr)
library(magrittr)
library(Hmisc)
library(colorspace)
library(ade4)
library(vegan)
library(gridExtra)

source("ScreePlot.R")
source("CIAPlots.R")

pathtoplots <- file.path(getwd(), "Plots/")#change me 
  
```
# MOMS-PI

The MOMS-PI data can be loaded as follows.

## 16S data

Load 16S data as a matrix, rows are Greengene IDs, columns are sample names:

```{r, echo = TRUE}
data("momspi16S_mtx")

```

Load the Greengenes taxonomy table as a matrix, rows are Greengene IDs, columns are taxonomic ranks:

```{r, echo = TRUE}
data("momspi16S_tax")
# Check if Greengene IDs match between the 16S and taxonomy data
# all.equal(rownames(momspi16S_mtx), rownames(momspi16S_tax)) # Should be TRUE
```

Load the 16S sample annotation data as a matrix, rows are samples, columns are annotations:

```{r, echo = TRUE}
data("momspi16S_samp")

# Check if sample names match between the 16S and sample data
# all.equal(colnames(momspi16S_mtx), rownames(momspi16S_samp)) # Should be TRUE
```

The `momspi16S` function assembles those matrices into a `phyloseq` object.

```{r, message=FALSE}
momspi16S_phyloseq <- momspi16S()
momspi16S_phyloseq
```

## Cytokine data

The MOMS-PI cytokine data can be loaded as a matrix, rownames are cytokine names, colnames are sample names:

```{r}
data("momspiCyto_mtx")
dim(momspiCyto_mtx)
```

Load the cytokine sample annotation data as a matrix, rows are samples, columns are annotations:

```{r}
data("momspiCyto_samp")
dim(momspiCyto_samp)
# Check if sample names match between the 16S and sample data
# all.equal(colnames(momspiCyto_mtx), rownames(momspiCyto_samp)) # Should be TRUE
```

The function `momspiCytokines` will make a `SummarizedExperiment` containing cytokine data

```{r}
momspiCyto <- momspiCytokines()
momspiCyto
```

The cytokine data contains data for `r nrow(momspiCyto)` cytokines over `r ncol(momspiCyto)` samples.

# Multi-table analysis

Combine 16S and cytokines data

```{r, echo = FALSE}
#order both sets by visit number within a subject
momspi16S_samp <- momspi16S_samp[
  with(momspi16S_samp, order(subject_id, sample_body_site, visit_number)),
] 

momspiCyto_samp <- momspiCyto_samp[
  with(momspiCyto_samp, order(subject_id, sample_body_site, visit_number)),
] 

#merge data by subject id and sample_body_site
combined_samp <- merge(momspi16S_samp, momspiCyto_samp, 
                       by = c("subject_id", "sample_body_site", 
                        "project_name", "study_full_name",
                        "subject_gender", "subject_race"))

```

```{r}
#select data collected at the same visit
combined_samp <- merge(momspi16S_samp, momspiCyto_samp, 
                       by = c("subject_id", "sample_body_site", 
                        "project_name", "study_full_name",
                        "subject_gender", "subject_race",
                        "visit_number"))

table(combined_samp$visit_number)

```

Select first visit data, this assures that samples we look at were taken at the same time and at the first or second trimester. We don't have trimesters information in the data, but know it from the study design. 

```{r, echo=FALSE}
#select earliest visit from each subject
#combined_samp.agg <- aggregate(visit_number ~  subject_id, data = combined_samp, min)
#combined_samp <- merge(combined_samp.agg, combined_samp, by = c("subject_id"))
#combined_samp <- combined_samp[combined_samp$visit_number %in% c(1,2),]
```


```{r, echo = TRUE}
#select data from first visit only
combined_samp <- combined_samp[combined_samp$visit_number ==  1,]

table(combined_samp$sample_body_site)#all vaginal samples
```

The two objects we use for combined 16S and cytokines analysis are: 'combined_16S_mtx' and 'combined_Cyto_mtx'. Phylogenetic information for those OTUs is available in 'tax_table(combined_16S_phyloseq)' object. 

```{r}
#select 16S data for those samples
combined_16S_phyloseq <- subset_samples(momspi16S_phyloseq, file %in% combined_samp$file.x)

#get rif of otus that are not observed in any sample for this subset
combined_16S_phyloseq %<>%
    taxa_sums() %>%
    is_greater_than(0) %>%
    prune_taxa(combined_16S_phyloseq)

combined_16S_mtx <- otu_table(combined_16S_phyloseq)

combined_Cyto_mtx <- momspiCyto_mtx[, colnames(momspiCyto_mtx) %in% combined_samp$file.y ]
dim(combined_Cyto_mtx)

```

We match the samples (contained in columns of both tables)  by the file names contained in colnames of each table.

In 'combined_samp' object the names of matched files names for 16S data are recorded in column 'file.x' and for cytokines data in column 'file.y'. 

```{r}
#make sure all samples across 3 tables are in the same order
combined_samp <- combined_samp[order(combined_samp$subject_id),]
#reorder cytokines samples
combined_Cyto_mtx <- combined_Cyto_mtx[,combined_samp$file.y]
#reorder taxa samples
combined_16S_mtx <- combined_16S_mtx[,combined_samp$file.x]
```

# Co-inertia analysis

## Basics:

\begin{itemize}

	\item Let $Z_1$ and  $Z_2$ be 16S and cytokines tables respectively
	\item rows: same $n$ women at first visit 
	\item Columns: $p_1$ taxa, $p_2$ cytokines  
	\item To visualize differences/similarities among taxa and cytokines data sets we view samples as  rows while taxa and cytokines as columns
	\item PCA analysis for each table: $(X, Q_X, D)$ and $(Y, Q_Y, D)$
	\item Co-inertia axes: $Y^TDX = K\Lambda^{1/2}A^T$ of decomposition $(Y^TDX,Q_X, Q_Y)$
	\item Plot $F_X = XA $ and $F_Y = YK$
\end{itemize}

```{r}
combined_16S_mtx <- t(combined_16S_mtx)
combined_16S_mtx <- combined_16S_mtx/apply(combined_16S_mtx, 1, sum)
combined_Cyto_mtx <- t(combined_Cyto_mtx)

#cut the last 5 characters that correspond to the -omics type identifier
rownames(combined_Cyto_mtx) <- substr(
  rownames(combined_Cyto_mtx), 1,nchar(rownames(combined_Cyto_mtx))-5)

rownames(combined_16S_mtx) <- substr(
  rownames(combined_16S_mtx), 1,nchar(rownames(combined_16S_mtx))-5)

#make sure all rownames match
all(rownames(combined_16S_mtx) == rownames(combined_16S_mtx))
```

We first center  16S data to work with PCA on the covariance matrix $\Sigma_X = Cov(X)$ and  then, to normalize the data, we divide each value of $X$ by the total variance: $\sqrt{\mbox{tr}(\Sigma_X)}$, which is equivalent to dividing the matrix by $\sqrt{\sum_{k=1}^r \lambda_k}$, where $\lambda_k$ are the eigevalues of $\Sigma_X$ and $r$ is the rank of $X$. This is the standartization approach used in multiple co-inertia analysis, which combines several tables. 

```{r}
taxa_mtx <- scale(combined_16S_mtx, center = TRUE, scale = FALSE)
#use fast trace computation formula: tr(A^B) = sum(A*B), where '*' operator refers to elemetwise product
taxa_tr <- sum(taxa_mtx*taxa_mtx)/(dim(taxa_mtx)[1]-1)
taxa_mtx <- taxa_mtx/sqrt(taxa_tr)
taxa.pca <- dudi.pca(taxa_mtx, scannf=FALSE, nf =61,
                     center = FALSE, scale = FALSE)
breaks <- c(50, 70, 80, 90)
scree.taxa.pca <- screePlot(taxa.pca, breaks) + 
           theme( panel.background = element_rect(fill = "white"),
                  panel.grid.major = element_line(colour = "grey90"))
```

Cytokines PCA on centered and scaled data, also normalized by the square root of total variances. 

```{r}
cyto_mtx <- scale(combined_Cyto_mtx, center = TRUE, scale = TRUE)
cyto_tr <- sum(cyto_mtx*cyto_mtx)/(dim(cyto_mtx)[1]-1)
cyto_mtx <- cyto_mtx/sqrt(cyto_tr)
cyto.pca <- dudi.pca(cyto_mtx, scannf=FALSE, nf =61,
                     center = FALSE, scale = FALSE)
breaks <- c(50, 70, 80, 90)
scree.cyto.pca <- screePlot(cyto.pca, breaks) + 
           theme( panel.background = element_rect(fill = "white"),
                  panel.grid.major = element_line(colour = "grey90"))

```

```{r, echo = FALSE}
#extract legend
#https://github.com/hadley/ggplot2/wiki/Share-a-legend-between-two-ggplot2-graphs
g_legend<-function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)}

mylegend<-g_legend(scree.cyto.pca)

scree.pca <- grid.arrange(arrangeGrob(scree.taxa.pca + theme(legend.position="none") + ggtitle("Taxa PCA"),
                         scree.cyto.pca + theme(legend.position="none")+ ggtitle("Cytokines PCA"),
                         nrow=1),
             mylegend, nrow=2,heights=c(10, 1))
scree.pca
```

```{r, echo = FALSE}
p.taxa_pca <- PlotPCA(Res.pca = taxa.pca,  Env_Var = data.frame(groups), 
                     axes = c(1,2), 
                     color = NULL,
                     shape= NULL,
                     plot_type = "samples",
                     ptColor = "blue")+
              theme_bw() + geom_point(color = "red")
```

```{r, echo = FALSE}
p.cyto_pca <- PlotPCA(Res.pca = cyto.pca,  Env_Var = data.frame(groups), 
                     axes = c(1,2), 
                     color = NULL,
                     shape= NULL,
                     plot_type = "samples",
                     ptColor = "blue")+
              theme_bw() + geom_point(color = "blue")
```


```{r, echo=FALSE}
grid.arrange(p.taxa_pca+ggtitle("taxa samples PCA"), p.cyto_pca+ggtitle("cytokines samples PCA"), ncol = 2)
```



```{r, echo = FALSE}
p.taxa_pca.sp <- PlotPCA(Res.pca = taxa.pca,  Env_Var = data.frame(groups), 
                     axes = c(1,2), 
                     color = NULL,
                     shape= NULL,
                     plot_type = "species",
                     ptColor = "red")+
              theme_bw() + geom_point(color = "red")
```

```{r, echo = FALSE}
p.cyto_pca.sp <- PlotPCA(Res.pca = cyto.pca,  Env_Var = data.frame(groups), 
                     axes = c(1,2), 
                     color = NULL,
                     shape= NULL,
                     plot_type = "species",
                     ptColor = "red")+
              theme_bw() + geom_point(color = "blue")
```


```{r, echo=FALSE}
grid.arrange(p.taxa_pca.sp+ggtitle("taxa variables PCA"), p.cyto_pca.sp+ggtitle("cytokines variables PCA"), ncol = 2)
```



```{r, echo=FALSE, eval = FALSE}
# Individual tables analysis PCoA
#calculate pairwise distances
cyto_dist <- vegdist(combined_Cyto_mtx, method= "bray", binary_dist=F)

n.pc = 3
cyto_mds <- dudi.pco(cyto_dist, scannf = FALSE, nf = n.pc)
#Samples Plots $l1
p.cyto_mds <- PlotPCA(Res.pca = cyto_mds,  Env_Var = data.frame(groups), 
                     axes = c(1,2), 
                     color = NULL,
                     shape= NULL,
                     plot_type = "samples",
                     ptColor = "blue")+
              theme_bw() + geom_point(color = "blue")
             
```

```{r, echo=FALSE, eval = FALSE}
#calculate pairwise distances
taxa_dist <- vegdist(combined_16S_mtx, method= "bray", binary_dist=F)

n.pc = 3
taxa_mds <- dudi.pco(taxa_dist, scannf = FALSE, nf = n.pc)
#Samples Plots $l1
p.taxa_mds <- PlotPCA(Res.pca = taxa_mds,  Env_Var = data.frame(groups), 
                     axes = c(1,2), 
                     color = NULL,
                     shape= NULL,
                     plot_type = "samples",
                     ptColor = "blue")+
              theme_bw() + geom_point(color = "red")
```

```{r, echo=FALSE, eval = FALSE}
grid.arrange(p.taxa_mds+ggtitle("taxa PCoA"), p.cyto_mds+ggtitle("cytokines PCoA"), ncol = 2)
```


#Combine the tables using co-inertia

```{r}
coin <- coinertia(taxa.pca, cyto.pca, scannf = FALSE, nf = 2)
```

RV coefficient -- measure of similarity between 16S and cytokines tables

```{r}
RV<- coin$RV
RV
```

Sample scores plots. Length of the arrows indicates the samples that have larger differences across two data sets.

```{r}

Samp.coin <- CoinertiaPlot(coin = coin, 
                  Quant = 0.9, Prop.Var = 0.9,
                  Env_Var=NULL, 
                  Env_Var2 = NULL,
                  color=NULL, shape=NULL,
                  PtColor= "magenta",PtShape=1, PtSize=1,
                  linetype=1, LblSize=3, 
                  LabelsOpt = NULL,
                  ArrLen=0.10, ArrAngle=20)

Samp.coin$p + theme_bw()

```

Samples with largest difference across two data sets. Samples with arrow lengths in 0.9 quatile are chosen. 

```{r}
#Taxa with major differences across two sets
rownames(Samp.coin$Dissimilarity[Samp.coin$Dissimilarity$Quantile >= 0.9, ])
```

Plots of  taxa weights 

```{r, eval = FALSE}
#Plot Canonical weights based on No BV taxa as reference point
CW.coin.ca <- PlotCW(coin = coin, name="Coin", path = pathtoplots, color = "magenta",
            Title1 = "Canonical weights for taxa",
            Title2 = "Canonical weights for cytokines",
            Labels1 = colnames(combined_16S_mtx), Labels2 = colnames(combined_Cyto_mtx),
            scale = TRUE,
            PtShape=2, PtSize=2,linesize=0.4,
            linetype=1, LblSize=2.5, 
            ArrLen=0.15, ArrAngle=10,
            TitleSize = 10) 

```

```{r}
p.vars <- PlotCoinVars(coin, tab1 = "taxa", tab2 = "cytokines", 
             Labels1 = NULL, #colnames(combined_16S_mtx)
             Labels2 = colnames(combined_Cyto_mtx),
             label = TRUE,PtSize=2,LblSize=2,
             hjust = 0, vjust = -1.5)
p.vars
```

Taxa that correspond to larger lodings: 324, 53, 263, 542, 290, 494, 57, 60

```{r, echo = TRUE}
taxa.inx <- c(324, 53, 263, 542, 290, 494, 57, 60)
taxa.ids <- colnames(combined_16S_mtx)[taxa.inx]
#look up these taxa ids in phyloqes
tax_table(momspi16S_phyloseq)[rownames(tax_table(momspi16S_phyloseq)) %in% taxa.ids,c("Genus", "Species")]
```

 

## Additional details about the method.

\begin{itemize}
\item $Z_{n \times p}$ data matrix
 	\begin{itemize}
		\item $n$ samples
		\item $p$ variables (taxa)
	\end{itemize}
\item Transformation $Z_{n \times p} \rightarrow X_{n \times p}$
 		\begin{itemize}
		\item Different transformation for each ordination method   
		\item Column weights: $Q_{p \times p}$
		\item Row weights: $D_{n \times n}$
	\end{itemize}
\item Statistical triplet: $(X, Q, D)$
\item Eigen decomposition of $(WD)_{n \times n} =  XQX^TD$ or $(VQ)_{p \times p} =  X^TDXQ$
\item Write: $X = K\Lambda^{1/2}A^T$
	\begin{itemize}
		\item $K^TDK = \mathbb I_r$, $K$ - samples scores
		\item $A^TQA = \mathbb I_r$, $A$ - taxa loadings
		\item $r$ rank of $X$
	\end{itemize}
\end{itemize}


```{r, echo = FALSE, out.width = '100%', out.height= '100%', results='hold'}
knitr::include_graphics(file.path(getwd(), "Plots/TriplDiagFull.pdf"))
```

Given statistical triplet $(X, Q, D)$   the PCA is 
\begin{itemize}
\item  $X_{n \times p}$ - \textcolor{blue}{column centered and scaled} matrix
	\begin{itemize}
		\item $n$ number of samples
		\item $p$ number of taxa
	\end{itemize}
\item   $Q = \mathbb{I}_p$ 
\item  Correlation PCA: $Q = diag \{sd(\boldsymbol z_1), \dots, sd(\boldsymbol z_p)\}$, where $sd(\boldsymbol z_j)$ standard deviation for  $j^{th}$ column of original data matrix $Z$ 
\item  $D = diag\{\frac{1}{n}, \dots, \frac{1}{n}\} = \frac{1}{n}\mathbb{I}_n$ 
\item  Then  eigendecomposition of 
$$
 \textcolor{blue}{VQ} = X^TDXQ = \frac{1}{n}X^TX = \textcolor{blue}{\Sigma}
$$ 
\item Inertia: $\textcolor{blue}{\mathcal I} = tr(\Sigma) = \textcolor{blue}{\sum_{i=1}^p \sigma^2_i} = \sum_{i=1}^r \lambda_i$
\end{itemize}

Co-inertia analysis idea:
	\begin{itemize}
		\item Analysis of $Z_1, Z_2, \dots, Z_K$. Same rows, but different columns
		\item  $Z_1$ -- 16S data 
		\item  $Z_2$ -- cytokines data 
  \end{itemize}


\begin{enumerate}
\item Co-Inertia Analysis
	\begin{itemize}
		\item Seeks  a \textcolor{blue}{common structure} between two datasets $Z_1$ and $Z_2$
		\item Two tables: $Z_1$ and $Z_2$ of dimensions $n \times p_1$ and $n \times p_2$
		\item Use any one table method to obtain triplets: $(X, Q_X, D)$ and $(Y, Q_Y, D)$
		\item \textcolor{blue}{coinertia} -- coinertia analysis function in ade4
		\item Co-inertia triplet: $(Y^TDX, Q_X, Q_Y)$
		\item Decomposition: $Y^TDX = K\Lambda^{1/2}A^T$
		\item Project $X$ and $Y$ onto $A$ and $K$ respectively
		\item Let $A = XQ_XX^TD$ and $B = YQ_YY^TD$
		\item RV coefficient: $RV(A,B) = \frac{tr(AB^T)}{\sqrt{tr(AA^T)tr(BB^T)}} = \frac{\mathcal I_{XY}}{\sqrt{\mathcal I_X} \sqrt{\mathcal I_Y}}$
		\item $RV \in [0,1]$  with values closer to $1$ indicating stronger  similarity among $X$ and $Y$
	\end{itemize}
\end{enumerate}

Objects in ade4

SVD of a transformed table $X =  K\Lambda^{1/2}A^T$

$k$ -- rank of $X$

\begin{flushleft}
\resizebox{1\textwidth}{!}{\begin{minipage}{\textwidth}
\begin{tabular}{lcll} \hline
matrix 			&	dimensions	&ade4	&description \\ \hline
$Z$				&	$n \times p$	&		&raw data; argument for dudi functions \\
$X$				&	$n \times p$	&\$tab	&transformed data 				\\
$Q$				&	$p \times p$	&\$cw	&diagonal matrix of column weights 	\\
$D$				&	$n \times n$	&\$lw	&diagonal matrix of row weights 	\\
$A$				&	$p \times k$	&\$c1	&the principal axes or column normed scores \\
$K$				&	$n \times k$	&\$l1		&the row normed scores					\\
$\Lambda$		&	$k\times k$	&\$eig	&diagonal matrix of the $k$ largest eigenvalues \\
$A\Lambda^{1/2}$	&	$p \times k$	&\$co	&column coordinates						\\
$K\Lambda^{1/2}$	&	$n \times k$	&\$li		&the row coordinates					\\
\end{tabular} 
  \end{minipage}}\\
\end{flushleft}

```{r}

```

